A Gaussian process model of human e\-lect\-ro\-cort\-i\-co\-graph\-ic data

### 2020-03-09 paper/main.tex

\begin{figure}
  \centering \includegraphics[width=\textwidth]{figs/methods}
  \caption{\textbf{Methods overview.}  \textbf{A.~Electrode locations.}  Each
  dot reflects the location of a single electrode implanted in the brain of a
  Dataset 1 patient.  A held-out recording location from one patient is
  indicated in red, and the patient's remaining electrodes are indicated in
  black. The electrodes from the remaining patients are colored by $k$-means
  cluster (computed using the full-brain correlation model shown in Panel D).
  \textbf{B.~Radial basis function kernel.} Each electrode contributed by the
  patient (black) weights on the full set of locations under consideration (all
  dots in Panel A, defined as $\bar{R}$ in the text).  The weights fall off with
  positional distance~\citep[in MNI152 space;][]{LyttEtal07} according to an RBF. \textbf{C.~Per-patient
  correlation matrices.}  After computing the pairwise correlations between the
  recordings from each patient's electrodes, we use RBF-weighted averages to
  estimate correlations between all locations in $\bar{R}$.  We obtain an
  estimated full-brain correlation matrix using each patient's data.
  \textbf{D.~Merged correlation model.}  We combine the per-patient correlation
  matrices (Panel C) to obtain a single full-brain correlation model that
  captures information contributed by every patient.  Here we have sorted the
  rows and columns to reflect $k$-means clustering labels~\citep[using
  $k$=7;][]{YeoEtal11}, whereby we grouped locations based on their correlations
  with the rest of the brain (i.e., rows of the matrix displayed in the panel).
  The boundaries denote the cluster groups.  The rows and columns of Panel C
  have been sorted using the Panel D-derived cluster labels.
  \textbf{E.~Reconstructing activity throughout the brain.}  Given the observed
  recordings from the given patient (shown in black; held-out recording is shown
  in blue), along with a full-brain correlation model (Panel D), we use
  Equation~\ref{eqn:reconstruction} to reconstruct the most probable activity at
  the held-out location (red).} \label{fig:methods}
\end{figure}

We first applied fourth order Butterworth notch filters to remove 60 Hz ($\pm$
.5 Hz) line noise from every recording (from every electrode). Next, we
downsampled the recordings (regardless of the original samplerate) to 250 Hz.
This downsampling step served to both normalize for differences in sampling
rates across patients and to ease the computational burden of our subsequent
analyses.  We then excluded any electrodes that showed putative epileptiform
activity. Specifically, we excluded from further analysis any electrode that
exhibited a maximum kurtosis of 10 or greater across all of that patient's
recording sessions.  We also excluded any patients with fewer than 2 electrodes
that passed this criteria, as the SuperEEG algorithm requires measuring
correlations between 2 or more electrodes from each patient.  For Dataset 1,
this yielded clean recordings from 4168 electrodes implanted throughout the
brains of 67 patients (Fig.~\ref{fig:methods}A, colored dots); for Dataset 2,
this yielded clean recordings from 5023 electrodes implanted in 78 patients.
Each individual patient contributed electrodes from a limited set of brain
locations, which we localized in a common space~\citep[MNI152;][]{GrabEtal06};
an example Dataset 1 patient's 54 electrodes that survived the kurtosis
thresholding procedure are highlighted in black and red
(Fig.~\ref{fig:methods}A).

@@

Although both datasets we examined provide good full-brain coverage (when
considering data from every patient), electrodes were not sampled uniformly
throughout the brain.  For example, in our patient population, electrodes are
more likely to be implanted in regions like the medial temporal lobe (MTL), and
are rarely implanted in occipital cortex (Fig.~\ref{fig:density}A, B).
Separately for each dataset, for each voxel in the 4~mm$^3$ voxel MNI152 brain,
we computed the proportion of electrodes in the dataset that were contained
within a 20 MNI unit radius sphere centered on that voxel.  We defined the
\textit{density} at that location as this proportion. Across Datasets 1 and 2,
the electrode placement densities were similar (correlation by voxel: $r = 0.6,
p < 10^{-10}$).  We wondered whether regions with good covererage might be
associated with better reconstruction accuracy. For example,
Figures~\ref{fig:corrmap}A and B indicate that many electrodes in the MTL (which
tends to be relatively densely sampled) have relatively high reconstruction
accuracy, and occipital electrodes (which tends to be relatively sparsely
sampled) tend to have relatively low reconstruction accuracy.  To test whether
this held more generally across the entire brain, for each dataset we computed
the electrode placement density for each electrode from each patient (using the
proportion of \textit{other} patients' electrodes within 20 MNI units of the
given electrode).  We then correlated these density values with the
across-patient reconstruction accuracies for each electrode.  We found no
reliable correlation between reconstruction accuracy and density for Dataset 1
($r = 0.05, p = 0.70$) and a reliable negative correlation for Dataset 2 ($r =
-0.21, p = 0.05$).  This indicates that the reconstruction accuracies we
observed are \textit{not} driven solely by sampling density, but rather may also
reflect higher order properties of neural dynamics such as functional
correlations between distant voxels~\citep{BetzEtal17b}.

Prior work in humans and animals has shown that the spatial profile of the local
field potential differs by frequency band~\citep[e.g., with respect to volume
conductance properties and contribution to the local field
potential;][]{BuzsEtal12, FrieEtal07, CronEtal11}.  For example, lower frequency
components of the local field potential tend to have higher power and extend
further in space than high-frequency components~\citep[e.g., ][]{MillEtal07,
MannEtal09}.  We wondered whether the reconstructions we observed might be
differently weighting or considering the contributions of activity at different
frequency bands.  We therefore examined a range of frequency bands ($\delta$:
2--4 Hz; $\theta$: 4--8 Hz; $\alpha$: 8--12 Hz; $\beta$: 12--30 Hz; $\gamma_L$:
30--60 Hz; and $\gamma_H$: 60--100 Hz), along with a measure of broadband power.
We used second-order Butterworth bandpass filters to compute the activity
patterns within each narrow frequency band.  We defined broadband power as the
mean height of a linear robust regression fit in log-log space to the order 4
Morelet wavelet-computed power spectrum at 50 log-spaced frequencies from from
2--100 Hz~\citep{MannEtal09}. We then repeated our within-subject and
across-subject cross-validated reconstruction accuracy tests (analogous to
Fig.~\ref{fig:corrmap}) separately for each frequency band
(Fig.~\ref{fig:frequency}). (We also carried out a similar analysis on the
Hilbert transform-computed spectral power within each narrow band; see
Fig.~\freqpower.) Across both datasets, we found that our approach is best at
reconstructing patterns of broadband activity (right-most bars in
Figs.~\ref{fig:frequency}A and D), a correlate of population firing
rate~\citep{MannEtal09}.  We also achieved good reconstruction accuracy within
each narrow frequency band (Figs.~\ref{fig:frequency} and \freqpower).  Activity
at lower frequencies ($\delta$, $\theta$, $\alpha$, and $\beta$) tended to be
reconstructed better than high-frequency patterns ($\gamma_L$ and $\gamma_H$),
with reconstruction accuracy peaking in the $\alpha$ band.  Overall, these
results indicate that our approach is able to accurately recover information
within the 2--100 Hz range.

@@

\begin{figure}
  \centering \includegraphics[width=\textwidth]{figs/networks}
  \caption{\textbf{Most informative recording locations by frequency band.}
  \textbf{A.~Intersections between information score maps by frequency band.}
  The regions indicated in each row depict the intersection between the top 10\%
  most informative locations across Datasets 1 and 2.  \textbf{B.~Network
  memberships of the most informative brain regions.}  The pie charts display
  the proportions of voxels in each region that belong to the seven networks
  identified by \cite{YeoEtal11}.  The relative sizes of the charts for each
  frequency band reflct the average across-subject reconstruction accuracies
  (Figs.~\ref{fig:corrmap}A, D).  The voxels in Panel A are colored according to
  the same network memberships.  \textbf{C.~Neurosynth terms associated with the
  most informative brain regions, by frequency band.}  The lists in each row
  display the top five neurosynth terms~\citep{RubiEtal17} decoded for each
  region.  \textbf{D.~Network parcellation map and legend.}  The parcellation
  defined by \cite{YeoEtal11} is displayed on the inflated brain maps.  The
  colors and network labels serve as a legend for Panels A and B.
  \textbf{E.~Combined map of the most informative brain regions.}  The map
  displays the union of the most informative maps in Panel A, colored by
  frequency band.  The labels also serve as a legend for Panel C.}
  \label{fig:networks}
\end{figure}

>>>>

\begin{figure}
  \centering \includegraphics[width=\textwidth]{figs/methods}
  \caption{\textbf{Methods overview.}  \textbf{A.~Electrode locations.}  Each
  dot reflects the location of a single electrode implanted in the brain of a
  Dataset 1 patient.  A held-out recording location from one patient is
  indicated in red, and the patient's remaining electrodes are indicated in
  black. The electrodes from the remaining patients are colored by $k$-means
  cluster (computed using the full-brain correlation model shown in Panel D).
  \textbf{B.~Radial basis function kernel.} Each electrode contributed by the
  patient (black) weights on the full set of locations under consideration (all
  dots in Panel A, defined as $\bar{R}$ in the text).  The weights fall off with
  positional distance (in MNI152 space) according to an RBF. \textbf{C.~Per-patient
  correlation matrices.}  After computing the pairwise correlations between the
  recordings from each patient's electrodes, we use RBF-weighted averages to
  estimate correlations between all locations in $\bar{R}$.  We obtain an
  estimated full-brain correlation matrix using each patient's data.
  \textbf{D.~Merged correlation model.}  We combine the per-patient correlation
  matrices (Panel C) to obtain a single full-brain correlation model that
  captures information contributed by every patient.  Here we have sorted the
  rows and columns to reflect $k$-means clustering labels~\citep[using
  $k$=7;][]{YeoEtal11}, whereby we grouped locations based on their correlations
  with the rest of the brain (i.e., rows of the matrix displayed in the panel).
  The boundaries denote the cluster groups.  The rows and columns of Panel C
  have been sorted using the Panel D-derived cluster labels.
  \textbf{E.~Reconstructing activity throughout the brain.}  Given the observed
  recordings from the given patient (shown in black; held-out recording is shown
  in blue), along with a full-brain correlation model (Panel D), we use
  Equation~\ref{eqn:reconstruction} to reconstruct the most probable activity at
  the held-out location (red).} \label{fig:methods}
\end{figure}

We first applied fourth order Butterworth notch filters to remove 60 Hz ($\pm$
0.5 Hz) line noise from every recording (from every electrode). Next, we
downsampled the recordings (regardless of the original samplerate) to 250 Hz.
This downsampling step served to both normalize for differences in sampling
rates across patients and to ease the computational burden of our subsequent
analyses.  We then excluded any electrodes that showed putative epileptiform
activity. Specifically, we excluded from further analysis any electrode that
exhibited a maximum kurtosis of 10 or greater across all of that patient's
recording sessions.  We also excluded any patients with fewer than 2 electrodes
that passed this criteria, as the SuperEEG algorithm requires measuring
correlations between 2 or more electrodes from each patient.  For Dataset 1,
this yielded clean recordings from 4168 electrodes implanted throughout the
brains of 67 patients (Fig.~\ref{fig:methods}A, colored dots); for Dataset 2,
this yielded clean recordings from 5023 electrodes implanted in 78 patients.
Each individual patient contributed electrodes from a limited set of brain
locations, which we localized in a common space~\citep[MNI152;][]{GrabEtal06};
an example Dataset 1 patient's 54 electrodes that survived the kurtosis
thresholding procedure are highlighted in black and red
(Fig.~\ref{fig:methods}A).

@@

Although both datasets we examined provide good full-brain coverage (when
considering data from every patient), electrodes were not sampled uniformly
throughout the brain.  For example, in our patient population, electrodes are
more likely to be implanted in regions like the medial temporal lobe (MTL), and
are rarely implanted in occipital cortex (Fig.~\ref{fig:density}A, B).
Separately for each dataset, for each voxel in the 4~mm$^3$ voxel MNI152 brain,
we computed the proportion of electrodes in the dataset that were contained
within a 20 MNI unit radius sphere centered on that voxel.  We defined the
\textit{density} at that location as this proportion. Across Datasets 1 and 2,
the electrode placement densities were similar (correlation by voxel: $r = 0.6,
p < 10^{-10}$).  We wondered whether regions with good covererage might be
associated with better reconstruction accuracy. For example,
Figures~\ref{fig:corrmap}C and D indicate that some electrodes in the MTL (which
tends to be relatively densely sampled) have relatively high reconstruction
accuracy, and occipital electrodes (which tends to be relatively sparsely
sampled) tend to have relatively low reconstruction accuracy.  To test whether
this held more generally across the entire brain, for each dataset we computed
the electrode placement density for each electrode from each patient (using the
proportion of \textit{other} patients' electrodes within 20 MNI units of the
given electrode).  We then correlated these density values with the
across-patient reconstruction accuracies for each electrode.  We found no
reliable correlation between reconstruction accuracy and density for Dataset 1
($r = 0.05, p = 0.70$) and a reliable negative correlation for Dataset 2 ($r =
-0.21, p = 0.05$).  This suggests that the reconstruction accuracies we
observed are \textit{not} driven solely by sampling density, but rather may also
reflect higher order properties of neural dynamics such as functional
correlations between distant voxels~\citep{BetzEtal17b}.

Prior work in humans and animals has shown that the spatial profile of the local
field potential differs by frequency band~\citep[e.g., with respect to volume
conductance properties and contribution to the local field
potential;][]{BuzsEtal12, FrieEtal07, CronEtal11}.  For example, lower frequency
components of the local field potential tend to have higher power and extend
further in space than high-frequency components~\citep[e.g., ][]{MillEtal07,
MannEtal09}.  We wondered whether the reconstructions we observed might be
differently weighting or considering the contributions of activity at different
frequency bands.  We therefore examined a range of frequency bands ($\delta$:
2--4 Hz; $\theta$: 4--8 Hz; $\alpha$: 8--12 Hz; $\beta$: 12--30 Hz; $\gamma_L$:
30--60 Hz; and $\gamma_H$: 60--100 Hz), along with a measure of broadband (BB)
power. We used second-order Butterworth bandpass filters to compute the activity
patterns within each narrow frequency band.  We defined broadband power as the
mean height of a linear robust regression fit in log-log space to the order 4
Morelet wavelet-computed power spectrum at 50 log-spaced frequencies from from
2--100 Hz~\citep{MannEtal09}. We then repeated our within-subject and
across-subject cross-validated reconstruction accuracy tests (analogous to
Fig.~\ref{fig:corrmap}) separately for each frequency band
(Fig.~\ref{fig:frequency}). (We also carried out a similar analysis on the
Hilbert transform-computed spectral power within each narrow band; see
Fig.~\freqpower.) Across both datasets, we found that our approach is best at
reconstructing patterns of broadband activity (right-most bars in
Figs.~\ref{fig:frequency}A and D), a correlate of population firing
rate~\citep{MannEtal09}.  We also achieved good reconstruction accuracy within
each narrow frequency band (Figs.~\ref{fig:frequency} and \freqpower).  Activity
at lower frequencies ($\delta$, $\theta$, $\alpha$, and $\beta$) tended to be
reconstructed better than high-frequency patterns ($\gamma_L$ and $\gamma_H$),
with reconstruction accuracy peaking in the $\alpha$ band.  Overall, these
results indicate that our approach is able to accurately recover information
within the 2--100 Hz range.

@@

\begin{figure}
  \centering \includegraphics[width=\textwidth]{figs/networks}
  \caption{\textbf{Most informative recording locations by frequency band.}
  \textbf{A.~Intersections between information score maps by frequency band.}
  The regions indicated in each row depict the intersection between the top 10\%
  most informative locations across Datasets 1 and 2.  \textbf{B.~Network
  memberships of the most informative brain regions.}  The pie charts display
  the proportions of voxels in each region that belong to the seven networks
  identified by \cite{YeoEtal11}.  The relative sizes of the charts for each
  frequency band reflect the average across-subject reconstruction accuracies
  (Figs.~\ref{fig:frequency}A, D).  The voxels in Panel A are colored according to
  the same network memberships.  \textbf{C.~Neurosynth terms associated with the
  most informative brain regions, by frequency band.}  The lists in each row
  display the top five neurosynth terms~\citep{RubiEtal17} decoded for each
  region.  \textbf{D.~Network parcellation map and legend.}  The parcellation
  defined by \cite{YeoEtal11} is displayed on the inflated brain maps.  The
  colors and network labels serve as a legend for Panels A and B.
  \textbf{E.~Combined map of the most informative brain regions.}  The map
  displays the union of the most informative maps in Panel A, colored by
  frequency band.  The labels also serve as a legend for Panel C.}
  \label{fig:networks}
\end{figure}
